Fully Extrinsic Type Inference (DRAFT 2025-06-20)

### 2025-07-30 doc/extrinsic_type_inference_paper/paper.tex
begin updating model typing rules

\begin{definition}
  \label{def:valuing}
  \emph{valuing}
  \hfill
  \small
  \boxed{\Downarrow e}
  \nopad
  \begin{mathpar}
    \inferrule { 
      \TODO{...}
    } {
      \Downarrow e 
    } 
  \end{mathpar}
\end{definition}

@@

\begin{definition}
  \label{def:model_typing}
  \emph{model typing}
  \hfill
  \small
  \boxed{\delta \satisfies e \hastype \tau}
  \nopad
  \begin{mathpar}
    \inferrule { 
      e \rightsquigarrow e'
      \\
      \delta \satisfies e' \hastype \tau 
    } {
      \delta \satisfies e \hastype \tau 
    }

    \inferrule { 
      \alpha \slash \tau \in \delta
      \\
      \delta \satisfies e \hastype \tau
    } {
      \delta \satisfies e \hastype \alpha 
    }

    \inferrule { 
      \delta \ \alpha \slash \J{LFP[}\alpha \J{]} \tau \satisfies e \hastype \tau
    } {
      \delta \satisfies e \hastype \J{LFP[} \alpha \J{]} \tau
    }

    \inferrule { 
      \delta \satisfies e \hastype \tau_l
    } {
      \delta \satisfies e \hastype \tau_l \J{|} \tau_r
    }

    \inferrule { 
      \delta \satisfies e \hastype \tau_r
    } {
      \delta \satisfies e \hastype \tau_l \J{|} \tau_r
    }

    \inferrule { 
      \textbf{dom}(\delta') \subseteq \Theta
      \\
      \delta \oplus \delta' \satisfies \Delta 
      \\
      \delta \oplus \delta' \satisfies e \hastype \tau
    } {
      \delta \satisfies e \hastype \J{EXI[}\Theta\J{]}\Delta \J{:} \tau
    }

    \inferrule { 
      \delta \satisfies e \hastype \tau_l
      \\
      \delta \satisfies e \hastype \tau_r
    } {
      \delta \satisfies e \hastype \tau_l \J{\&} \tau_r
    }

    \inferrule { 
      \forall \delta' \qua
      \textbf{dom}(\delta') \subseteq \Theta \implies
      \delta \oplus \delta' \VDash \Delta \implies 
      \delta \oplus \delta' \satisfies e \hastype \tau
      \\\\
      \exists \delta' \qua 
      \textbf{dom}(\delta') \subseteq \Theta \up 
      \delta \oplus \delta' \satisfies \Delta
    } {
      \delta \satisfies e \hastype \J{ALL[}\Theta\J{]}\Delta \J{:} \tau
    }

    \inferrule { 
      \J{<}l \J{>} v \in g
      \\
      \delta \satisfies v \hastype \tau 
      \\
      \forall v' \qua \J{<} l \J{>} v' \in g \implies v' = v
    } {
      \delta \satisfies g \hastype \J{<} l \J{>} \tau
    }

    \inferrule { 
      \TODO{diff antec}
      \\\\
      p \Uparrow \eta 
      \\
      \delta \satisfies f \hastype \tau_l \J{->} \tau_r
    } {
      \delta \satisfies \J{[}p \J{=>} e \J{]}\ f \hastype \tau_l \typdiff \eta \J{->} \tau_r
    }

    \inferrule { 
      \TODO{needs separate inter antec}
      \\\\
      \forall \sigma \qua 
      \delta \VDash p[\sigma] \hastype \tau_l
      \implies
      \delta  \satisfies e[\sigma] \hastype \tau_r
    } {
      \delta \satisfies \J{[}p \J{=>} e \J{]}\ f\hastype \tau_l \J{->} \tau_r
    }

    \inferrule { 
      \TODO{inter antec left}
      \\\\
      \delta \satisfies e \hastype \tau_l \J{->} \tau
    } {
      \delta \satisfies e \hastype (\tau_l \J{\&} \tau_r) \J{->} \tau
    }

    \inferrule { 
      \TODO{inter antec right}
      \\\\
      \delta \satisfies e \hastype \tau_r \J{->} \tau
    } {
      \delta \satisfies e \hastype (\tau_l \J{\&} \tau_r) \J{->} \tau
    }

    \inferrule { 
      \TODO{all antec}
      \\\\
      \textbf{dom}(\delta') \subseteq \Theta 
      \\
      \delta \oplus \delta' \satisfies \Delta 
      \\
      \delta \oplus \delta' \satisfies e \hastype \tau_l \J{->} \tau_r
    } {
      \delta \satisfies e \hastype (\J{ALL[}\Theta \J{]}\Delta\ \tau_l) \J{->} \tau_r
    }

    \inferrule { 
      \TODO{union antec}
      \\\\
      \delta \satisfies e \hastype \tau_l \J{->} \tau
      \\
      \delta \satisfies e \hastype \tau_r \J{->} \tau
    } {
      \delta \satisfies e \hastype (\tau_l \J{|} \tau_r) \J{->} \tau
    }

    \inferrule { 
      \TODO{exi antec}
      \\\\
      \forall \delta' \qua 
      \textbf{dom}(\delta') \subseteq \Theta
      \implies
      \delta \oplus \delta' \VDash \Delta 
      \implies
      \delta \satisfies e \hastype \tau_l \J{->} \tau_r
    } {
      \delta \satisfies e \hastype (\J{EXI[} \Theta \J{]}\Delta\ \tau_l) \J{->} \tau_r
    }

    \inferrule { 
      \TODO{lfp antec}
      \\\\
      \alpha \notin \textbf{ftv}(\tau_l[\alpha\slash\J{@}]^+)
      \\
      \delta\ \alpha \slash (\J{LFP[}\alpha \J{]} \tau_l) \satisfies e \hastype \tau_l \J{->} \tau_r
    } {
      \delta \satisfies e \hastype (\J{LFP[}\alpha \J{]} \tau_l) \J{->} \tau_r
    }

    \inferrule { 
      \delta \satisfies e \hastype \tau_l
      \\
      \neg\ (\delta \satisfies e \hastype \tau_r)
    } {
      \delta \satisfies e \hastype \tau_l \typdiff \tau_r
    }

    \inferrule { 
    } {
      \delta \satisfies \J{@} \hastype \J{@}
    }

>>>>

\begin{definition}
  \label{def:valuing}
  \emph{valuing}
  \hfill
  \small
  \boxed{\downharpoonleft e}
  \nopad
  \begin{mathpar}
    \inferrule { 
      \TODO{...}
    } {
      \downharpoonleft 
      e 
    } 
  \end{mathpar}
\end{definition}

\begin{definition} 
  \label{def:type_sub_fold}
  Type Substitution Fold 
  \hfill
  \small
  \boxed{
    \tau [\alpha \slash \tau']^n
  }
  \\
  \[
  \begin{array}[t]{rclcl}
      \tau [\alpha \slash \tau']^0 
      &=& 
      \tau' 
      \\
      \tau [\alpha \slash \tau']^{(n + 1)} 
      &=& 
      \tau [\alpha \slash \tau[\alpha \slash \tau']^n]
  \end{array}
  \]
\end{definition}

@@

\TODO{completely redo model typing with well founded recursion}

\TODO{define monotonicity/antitonicity/polarity check}

\TODO{create recursive definition of least fixed point check}

\TODO{use separate typing for interpreting variables as finite types}

\TODO{use non-syntactic description of implication}

\TODO{remove implication antecedent rules}

\TODO{lift progression rule out to highest level for monotonic recursion}

\begin{definition}
  \label{def:model_core_typing}
  \emph{model core typing}
  \hfill
  \small
  \boxed{\delta \VDash e \hastype \tau}
  \nopad
  \begin{mathpar}
    \inferrule { 
      \alpha \wr^+ \tau
      \\
      \exists n \qua e \hastype \tau[\alpha \slash \J{BOT}]^n 
    } {
      \delta \VDash e \hastype \J{LFP[}\alpha\J{]}\tau 
    } 
  \end{mathpar}
\end{definition}

\begin{definition}
  \label{def:model_finite_typing}
  \emph{model finite typing}
  \hfill
  \small
  \boxed{e \hastype \tau}
  \nopad
  \begin{mathpar}
  \end{mathpar}
\end{definition}

\begin{definition}
  \label{def:model_typing}
  \emph{model typing}
  \hfill
  \small
  \boxed{\delta \satisfies e \hastype \tau}
  \nopad
  \begin{mathpar}
    % \inferrule { 
    %   e \rightsquigarrow e'
    %   \\
    %   \delta \satisfies e' \hastype \tau 
    % } {
    %   \delta \satisfies e \hastype \tau 
    % }

    % \inferrule { 
    %   \alpha \slash \tau \in \delta
    %   \\
    %   \delta \satisfies e \hastype \tau
    % } {
    %   \delta \satisfies e \hastype \alpha 
    % }

    % \inferrule { 
    %   \TODO{alt def}
    %   \\\\
    %   \forall \tau' \qua  
    %   \delta \satisfies \tau[\alpha\slash\tau'] \subtypes \tau'
    %   \implies
    %   \delta \satisfies e \hastype \tau'
    % } {
    %   \delta \satisfies e \hastype \J{LFP[} \alpha \J{]} \tau
    % }

    % \inferrule { 
    %   \delta \ \alpha \slash \J{LFP[}\alpha \J{]} \tau \satisfies e \hastype \tau
    % } {
    %   \delta \satisfies e \hastype \J{LFP[} \alpha \J{]} \tau
    % }

    % \inferrule { 
    %   \delta \satisfies e \hastype \tau_l
    % } {
    %   \delta \satisfies e \hastype \tau_l \J{|} \tau_r
    % }

    % \inferrule { 
    %   \delta \satisfies e \hastype \tau_r
    % } {
    %   \delta \satisfies e \hastype \tau_l \J{|} \tau_r
    % }

    % \inferrule { 
    %   \textbf{dom}(\delta') \subseteq \Theta
    %   \\
    %   \delta \oplus \delta' \satisfies \Delta 
    %   \\
    %   \delta \oplus \delta' \satisfies e \hastype \tau
    % } {
    %   \delta \satisfies e \hastype \J{EXI[}\Theta\J{]}\Delta \J{:} \tau
    % }

    % \inferrule { 
    %   \delta \satisfies e \hastype \tau_l
    %   \\
    %   \delta \satisfies e \hastype \tau_r
    % } {
    %   \delta \satisfies e \hastype \tau_l \J{\&} \tau_r
    % }

    % \inferrule { 
    %   \forall \delta' \qua
    %   \textbf{dom}(\delta') \subseteq \Theta \implies
    %   \delta \oplus \delta' \VDash \Delta \implies 
    %   \delta \oplus \delta' \satisfies e \hastype \tau
    %   \\\\
    %   \exists \delta' \qua 
    %   \textbf{dom}(\delta') \subseteq \Theta \up 
    %   \delta \oplus \delta' \satisfies \Delta
    % } {
    %   \delta \satisfies e \hastype \J{ALL[}\Theta\J{]}\Delta \J{:} \tau
    % }

    % \inferrule { 
    %   \J{<}l \J{>} v \in g
    %   \\
    %   \delta \satisfies v \hastype \tau 
    %   \\
    %   \forall v' \qua \J{<} l \J{>} v' \in g \implies v' = v
    % } {
    %   \delta \satisfies g \hastype \J{<} l \J{>} \tau
    % }

    % \inferrule { 
    %   \TODO{diff antec}
    %   \\\\
    %   p \Uparrow \eta 
    %   \\
    %   \delta \satisfies f \hastype \tau_l \J{->} \tau_r
    % } {
    %   \delta \satisfies \J{[}p \J{=>} e \J{]}\ f \hastype \tau_l \typdiff \eta \J{->} \tau_r
    % }

    % \inferrule { 
    %   \TODO{needs separate inter antec}
    %   \\\\
    %   \forall \sigma \qua 
    %   \delta \VDash p[\sigma] \hastype \tau_l
    %   \implies
    %   \delta  \satisfies e[\sigma] \hastype \tau_r
    % } {
    %   \delta \satisfies \J{[}p \J{=>} e \J{]}\ f\hastype \tau_l \J{->} \tau_r
    % }

    % \inferrule { 
    %   \TODO{inter antec left}
    %   \\\\
    %   \delta \satisfies e \hastype \tau_l \J{->} \tau
    % } {
    %   \delta \satisfies e \hastype (\tau_l \J{\&} \tau_r) \J{->} \tau
    % }

    % \inferrule { 
    %   \TODO{inter antec right}
    %   \\\\
    %   \delta \satisfies e \hastype \tau_r \J{->} \tau
    % } {
    %   \delta \satisfies e \hastype (\tau_l \J{\&} \tau_r) \J{->} \tau
    % }

    % \inferrule { 
    %   \TODO{all antec}
    %   \\\\
    %   \textbf{dom}(\delta') \subseteq \Theta 
    %   \\
    %   \delta \oplus \delta' \satisfies \Delta 
    %   \\
    %   \delta \oplus \delta' \satisfies e \hastype \tau_l \J{->} \tau_r
    % } {
    %   \delta \satisfies e \hastype (\J{ALL[}\Theta \J{]}\Delta\ \tau_l) \J{->} \tau_r
    % }

    % \inferrule { 
    %   \TODO{union antec}
    %   \\\\
    %   \delta \satisfies e \hastype \tau_l \J{->} \tau
    %   \\
    %   \delta \satisfies e \hastype \tau_r \J{->} \tau
    % } {
    %   \delta \satisfies e \hastype (\tau_l \J{|} \tau_r) \J{->} \tau
    % }

    % \inferrule { 
    %   \TODO{exi antec}
    %   \\\\
    %   \forall \delta' \qua 
    %   \textbf{dom}(\delta') \subseteq \Theta
    %   \implies
    %   \delta \oplus \delta' \VDash \Delta 
    %   \implies
    %   \delta \satisfies e \hastype \tau_l \J{->} \tau_r
    % } {
    %   \delta \satisfies e \hastype (\J{EXI[} \Theta \J{]}\Delta\ \tau_l) \J{->} \tau_r
    % }

    % \inferrule { 
    %   \TODO{lfp antec}
    %   \\\\
    %   \alpha \notin \textbf{ftv}(\tau_l[\alpha\slash\J{@}]^+)
    %   \\
    %   \delta\ \alpha \slash (\J{LFP[}\alpha \J{]} \tau_l) \satisfies e \hastype \tau_l \J{->} \tau_r
    % } {
    %   \delta \satisfies e \hastype (\J{LFP[}\alpha \J{]} \tau_l) \J{->} \tau_r
    % }

    % \inferrule { 
    %   \delta \satisfies e \hastype \tau_l
    %   \\
    %   \neg\ (\delta \satisfies e \hastype \tau_r)
    % } {
    %   \delta \satisfies e \hastype \tau_l \typdiff \tau_r
    % }

    % \inferrule { 
    % } {
    %   \delta \satisfies \J{@} \hastype \J{@}
    % }
